---
title: "p8123_final"
author: "Farizah Rob"
date: "2023-12-14"
output: pdf_document
---

```{r, message = F, warning = F}
library(tidyverse)
library(survey)
library(ggsurvey)
library(tableone)
library(gtsummary)
library(sjPlot)
```


```{r}
load("./data/nhanes_data.rda") #put files into data folder so everyone can run this
load("./data/nhanes_key.rda") 
load("./data/key_guide.rda")
```

**Problem statement:** Controlling blood pressure (BP) reduces the risk for cardiovascular disease. However, the prevalence of BP control (i.e., systolic BP < 140 and diastolic BP < 90) among US adults with hypertension has decreased since 2013. We invite teams to analyze publicly available data from US adults to help identify potential causes or correlates of worsening BP control among US adults with hypertension over the past decade, as this may allow for development of effective interventions to help control BP and prevent cardiovascular disease.


### Svydesign and subset for hypertension patients 2013 onwards

```{r}
dstrat <- svydesign(ids=~svy_psu, strata=~svy_strata, weights=~svy_weight_mec, data=nhanes_data, nest=TRUE) 

htsub <- subset(dstrat, svy_subpop_htn==1) #limited to hypertension patients

htsub2 <- subset(htsub, svy_year=="2013-2014" | svy_year=="2015-2016" | svy_year=="2017-2020") #limited to hypertension patents from 2013 onwards
```


### Some EDA

The **outcome** of interest -> **bp_control_140_90** OR **bp_uncontrolled_140_90**

```{r}
#age -> so for older people, BP control is definitely lower
age <- svyby(~factor(bp_control_140_90), ~factor(demo_age_cat), design=htsub2, svymean, na.rm=TRUE)
barplot(age, legend.text=c("BP uncontrolled", "BP controlled"), col=c("pink", "purple"), main="Proportion of BP control by age category") 

#gender -> not really any difference 
gender <- svyby(~factor(bp_control_140_90), ~factor(demo_gender), design=htsub2, svymean, na.rm=TRUE)
barplot(gender, legend.text=c("BP uncontrolled", "BP controlled"), col=c("pink", "purple"), main="Proportion of BP control by gender") #not much of a change

#race -> for Black people vs non-Black people, BP control is lower
race <- svyby(~factor(bp_control_140_90), ~factor(demo_race_black), design=htsub2, svymean, na.rm=TRUE)
barplot(race, 
        legend.text=c("BP uncontrolled", "BP controlled"), 
        names.arg=c("Non-Black", "Black"), 
        ylim=c(0, 1.0),
        args.legend = list(x = "topright"), 
        main="Proportion of BP control by race (non-Black vs Black)")

#diabetes
diabetes <- svyby(~factor(bp_control_140_90), ~factor(cc_diabetes), design=htsub2, svymean, na.rm=TRUE)
barplot(diabetes, 
        legend.text=c("BP uncontrolled", "BP controlled"), 
        names.arg=c("No Diabetes", "Diabetes"), 
        ylim=c(0, 1.0),
        args.legend = list(x = "topright"), 
        main="Proportion of BP control by diabetes")

#ckd
ckd <- svyby(~factor(bp_control_140_90), ~factor(cc_ckd), design=htsub2, svymean, na.rm=TRUE)
barplot(ckd, 
        legend.text=c("BP uncontrolled", "BP controlled"), 
        names.arg=c("No CKD", "CKD"), 
        ylim=c(0, 1.0),
        args.legend = list(x = "topright"), 
        main="Proportion of BP control by CKD")

#bmi 
bmi <- svyby(~factor(bp_control_140_90), ~factor(cc_bmi), design=htsub2, svymean, na.rm=TRUE)
barplot(bmi, 
        legend.text=c("BP uncontrolled", "BP controlled"), 
        ylim=c(0, 1.0),
        args.legend = list(x = "topright"), 
        main="Proportion of BP control by BMI")

#meds recommended -> 100% for those with uncontrolled BP
# look at med recommended and meds taken ratio by different subgroups?

table(nhanes_data$bp_med_recommended_jnc7, nhanes_data$bp_med_use) #48% of people recommended medication do not take medication

svyCreateCatTable(vars= c("bp_med_recommended_jnc7" ,"bp_med_use"), 
                  strata="demo_race", 
                  data=htsub2) #across all races, the % of med use is lower than % recommended -? this gap is highest for NH Black
```


```{r}
#for table 1 -> age category, gender, race, BMI, smoking status, diabetes, CKD, cholesterol, recommended medication use, medication use

# removed "bp_med_recommended_jnc7", "bp_med_recommended_accaha" for space

# didn't they say something in class about the Rao and Scott being wrong in R? not sure if we should include p-values here

# I don't think we should have p-values in the table 1 since we're not running models yet, just showing mean and standard deviation of the different covariates per outcome - Will

reset_gtsummary_theme()
theme_gtsummary_compact()
tab1 <- tbl_svysummary(data = htsub2, by = "bp_control_140_90", include = c("demo_age_cat", "demo_gender", "demo_race", "cc_bmi", "cc_ckd", "cc_diabetes", "cc_cvd_any", "bp_med_use", "chol_total", "chol_hdl", "chol_ldl", "chol_trig", "chol_med_use"),
                       label = list(demo_age_cat ~ "Age", 
                                    demo_gender ~ "Gender", 
                                    demo_race ~ "Race",
                                    cc_bmi ~ "BMI", 
                                    cc_ckd ~ "Chronic Kidney Disease", 
                                    cc_diabetes ~ "Diabetes", 
                                    cc_cvd_any ~ "Cardiovascular Disease",
                                    bp_med_use ~ "BP Medication Use", 
                                    chol_total ~ "Total Cholesterol", 
                                    chol_hdl ~ "HDL", 
                                    chol_ldl ~ "LDL", 
                                    chol_trig ~ "Triglycerides",
                                    chol_med_use ~ "Cholesterol Medication Use")) %>% add_p() %>%  add_stat_label() %>% modify_caption("**Subject Characteristics (Survey Weighted)**") %>% bold_labels() %>% 
  italicize_levels()

tab1


```

### Possible Figure 1:

I made a six-panel figure that could serve as our single Figure 1, feel free to change these plots to different things or remove ones we might not need. - Will

```{r}

par(mfrow = c(2, 3))

p1 <- svyhist(~bp_sys_mean, htsub2, main = "Average Systolic Blood Pressure NHANES 2013-2020", xlab = "Average Systolic Blood Pressure (mmHg)")

prop <- svymean(~bp_uncontrolled_140_90, htsub2, na.rm=TRUE)

p2 <- barplot(prop, 
        ylim=c(0, 1.0),
        main="Proportion of BP Control NHANES 2013-2020", names.arg = c("BP Uncontrolled", "BP Controlled"),
        ylab = "Proportion")

p3 <- svyplot(bp_sys_mean~demo_age_years, htsub2, style = "transparent", xlab = "Age (years)", ylab = "Average Systolic Blood Pressure (mmHg)", main = "Survey-Weighted Bubble Plot")

race <- svyby(~bp_uncontrolled_140_90, ~factor(demo_race), htsub2, svymean, na.rm=TRUE)

p4 <- barplot(race, 
        legend.text=c("BP uncontrolled", "BP controlled"), 
        ylim=c(0, 1.0),
        args.legend = list(x = "topright", inset = c(-0.5,-0.15), bty = "n"), 
        main="Proportion of BP Control by Race",
        ylab = "Proportion")

p5 <- svyboxplot(bp_sys_mean~factor(cc_smoke), htsub2, na.rm=TRUE, xlab = "Smoking Status", ylab = "Average Systolic Blood Pressure (mmHg)", main = "Systolic Blood Pressure per Smoking Status")

cvd <- svyby(~bp_uncontrolled_140_90, ~factor(cc_cvd_chd), htsub2, na.rm = T, svymean)

p6 <- barplot(cvd, 
        legend.text=c("BP uncontrolled", "BP controlled"), 
        ylim=c(0, 1.0),
        args.legend = list(x = "topright", inset = c(-0.5,-0.05), bty = "n"), 
        main="Proportion of BP Control by CVD History",
        xlab = "Cardiovascular Disease History",
        ylab = "Proportion")

```


### Regression models - really don't know which variables to select 

using the **bp_uncontrolled_140_90** variable instead of bp_control for regression

```{r}
m1 <- svyglm(bp_uncontrolled_140_90 ~ demo_age_cat + demo_race + demo_gender + cc_smoke + cc_bmi+ cc_diabetes + cc_ckd + cc_cvd_any + chol_trig + chol_ldl + chol_hdl, design = htsub2, family=quasibinomial()) #does it make sense to add bp medication use and recommended variables? they are prescribed bc of the uncontrolled bp

#need to set BMI reference value to normal level rather than underweight

summary(m1)

#need to look into subset of high cholesterol and hypertension?
```

### Modeling medication use with bp_uncontrolled_140_90 outcome

Model is limited to survey design subset of hypertension patients from 2013-2020. We selected medication variables concerning self-reported blood pressure medication use (yes/no), the number of antihypertensive medication pills taken per day, the use of vasodilators (yes/no), alpha and beta blockers (yes/no), the use of angiotensin receptor blockers, and the use of statins or other cholesterol medication for cholesterol control (yes/no). These variables were selected based on their prior association with blood pressure levels in existing clinical trials and research studies. 
```{r}

meds_binary <- svyglm(bp_uncontrolled_140_90 ~ bp_med_use + bp_med_n_pills + bp_med_vasod + bp_med_alpha + bp_med_beta + bp_med_angioten + chol_med_statin + chol_med_use, design = htsub2, family = quasibinomial())

summary(meds_binary)
```

Use of blood pressure medications, number of antihypertensive pills (1-4+), and angiotensin receptor blockers, were significantly associated with having uncontrolled blood pressure. The use of vasodilators, statins or other cholesterol medication, alpha and beta blockers, had non-statistically significant associations with having uncontrolled blood pressure.

### Modeling medication use with bp_sys_mean outcome

We now model the continuous outcome of the average systolic blood pressure with the same medication covariates among those with hypertension from 2013-2020. 

```{r}
meds_continuous <- svyglm(bp_sys_mean ~ bp_med_use + bp_med_n_pills + bp_med_vasod + bp_med_alpha + bp_med_beta + bp_med_angioten + chol_med_statin + chol_med_use, design = htsub2, family = gaussian())

summary(meds_continuous)
```

The use of blood pressure medications, number of antihypertensive pills (1), and angiotensin receptor blockers were significantly associated with mean systolic blood pressure levels in those with hypertension from 2013-2020. The use of (2-4+) antihypertensive pills per day, vasodilators, alpha and beta blockers, and statins and other cholesterol medications, had non-statistically significant associations with mean systolic blood pressure levels in those with hypertension from 2013-2020. 

## Nested Models
### Level One Demographic Characteristics.
```{r}
model_1 <- svyglm(bp_uncontrolled_140_90 ~ demo_age_cat + demo_race + demo_gender, #demographics
                  design = htsub2, family=quasibinomial())
summary(model_1)
```

### Level Two Health-related Characteristics 
```{r}
model_2 <- svyglm(bp_uncontrolled_140_90 ~ demo_age_cat + demo_race + demo_gender + #demographics
                    cc_smoke + cc_bmi+ cc_diabetes + cc_ckd + cc_cvd_any + chol_trig + chol_ldl + chol_hdl, #health-related
                  design = htsub2, family=quasibinomial())
summary(model_2)
```

### Level Three Medication Use
```{r}
model_3 <- svyglm(bp_uncontrolled_140_90 ~ demo_age_cat + demo_race + demo_gender + #demographics
                    cc_smoke + cc_bmi+ cc_diabetes + cc_ckd + cc_cvd_any + chol_trig + chol_ldl + chol_hdl + #health-related
                    bp_med_use + bp_med_n_pills + bp_med_vasod + bp_med_alpha + #medication
                    bp_med_beta + bp_med_angioten + chol_med_statin + chol_med_use, #medication
                  design = htsub2, family=quasibinomial())
summary(model_3)
```

I added the nested models based on three levels. Should we also make figures based on the significant variables from these models? If these models look good, I will proceed making the table two - Harry

### Forest plot of estimates from models 1-3

```{r}
axis.labels = c("Ages 45-64", "Ages 65-74", "Ages 75+", "Non-Hispanic Black", "Non-Hispanic Asian", "Hispanic", "Other Race", "Gender: Women", "Former Smoker", "Current Smoker", "BMI 25-30", "BMI 30-35", "BMI 35+", "Has Diabetes", "Has CKD", "Has CVD", "Triglycerides", "LDL Chol", "HDL Chol", "Uses BP Meds", "BP Meds per day: 1", "BP Meds per day: 2", "BP Meds per day: 3", "BP Meds per day: 4+", "Uses Vasodilators", "Uses Alpha-Blockers", "Uses Beta-Blockers", "Uses Angiotens", "Uses Statins", "Uses Cholesterol Meds")

axis.labels <- stringi::stri_reverse(axis.labels)

plot_models(model_1, model_2, model_3, transform = "exp", legend.title = "", show.p = T, grid = T, show.legend = F, m.labels = c("Model 1", "Model 2", "Model 3"))
```

